---
title: "Alaska Salmon 1922 - 2017"
format: html
author: Sofia Rodas
---

This quarto document contains the code to access the locally housed duckdb database that contains the data of Alaska salmon from 1922-2017. The data was originally downloaded from [The Knowledge Network for Biocomplexity (KNB)](https://knb.ecoinformatics.org/view/doi:10.5063/F1707ZTM). 

Code and instructions to make the relational database is located in the repository for this [project](https://github.com/sofiiir/eds213-finalproject.git). 


## Connecting to the local database

```{r}
# install packages
library(tidyverse)
library(DBI) # makes connections to the databases
library(duckdb)
library(dbplyr)
```

Connect to the local database using duckdb. 
```{r}
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "alaska-salmon-database/database.duckdb")

conn
```

```{r}
# view the tables contained in this database
DBI::dbListTables(conn)
```

## Look at the tables necessary to make the visual
```{r}
salmon_data <- tbl(conn, "Salmon")
head(salmon_data)
```

```{r}
locations <- tbl(conn, "Salmon_location")
head(locations)
```
# Overarching question

How many sampling are there for each species based on each sampling method throughout Alaska Salmon sampling from 1922 to 2017?
```{r}
# use dblyr code to run a query
sampled_salmon <- salmon_data |> 
  left_join(locations, by = "locationunique", keep = FALSE) |> 
  group_by(species, aslprojecttype) |> 
  summarise(num_sample = n()) |> 
  collect()

# view the output
sampled_salmon
```

# Visualize findings

There are a lot of different ASL Project Types. We'll just visualize the top three

```{r}
sampled_salmon |> 
  filter(aslprojecttype != "999") |> 
  group_by(aslprojecttype) |> 
  summarize(sum = sum(num_sample)) |> 
  arrange(desc(sum))
```
The top three ASL project type is commercial catch, escapement, and hatchery cost recovery.

```{r}

sampled_salmon |> 
  filter(aslprojecttype == "commercial catch" | aslprojecttype == "escapement" | aslprojecttype == "hatchery cost recovery") |> 
  ggplot() +
  geom_col(aes(x = species, y = num_sample)) +
  facet_wrap(vars(aslprojecttype)) + 
  coord_flip() +
  theme_minimal() + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  labs(x = "Salmon species",
       y = "Count")
```

```{r}
sampled_salmon |> 
  filter(aslprojecttype == "hatchery cost recovery") |> 
  ggplot() +
  geom_col(aes(x = species, y = num_sample)) +
  coord_flip() +
  theme_minimal() + 
  labs(x = "Salmon species",
       y = "Count",
       title = "Hatchery Cost Recovery")
```



## Close the database connection

```{r}
DBI::dbDisconnect(conn, shutdown = TRUE)
```

