---
title: "Alaska Salmon 1922 - 2017"
format: html
author: Sofia Rodas
---

This quarto document contains the code to access the locally housed duckdb database that contains the data of Alaska salmon from 1922-2017. The data was originally downloaded from [The Knowledge Network for Biocomplexity (KNB)](https://knb.ecoinformatics.org/view/doi:10.5063/F1707ZTM). 

Code and instructions to make the relational database is located in the repository for this [project](https://github.com/sofiiir/eds213-finalproject.git). 


## Connecting to the local database

```{r}
# install packages
library(tidyverse)
library(DBI) # makes connections to the databases
library(duckdb)
library(dbplyr)
```

Connect to the local database using duckdb. 
```{r}
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "alaska-salmon-database/database.duckdb")

conn
```

```{r}
# view the tables contained in this database
DBI::dbListTables(conn)
```

## Look at the tables necessary to make the visual
```{r}
salmon_data <- tbl(conn, "Salmon")
head(salmon_data)
```

```{r}
locations <- tbl(conn, "Salmon_location")
head(locations)
```
# Overarching question

How many sampling are there for each species based on each sampling method throughout Alaska Salmon sampling from 1922 to 2017?
```{r}
# use dblyr code to run a query assessing the number of salmon based on each project type
sampled_salmon <- salmon_data |> 
  left_join(locations, by = "locationunique", keep = FALSE) |> 
  group_by(species, aslprojecttype) |> 
  summarise(num_sample = n()) |> 
  collect()

# view the output
sampled_salmon
```

# Visualize findings

There are a lot of different ASL Project Types. We'll just visualize the top three!

First we remove the NA's for the project type. These missing project types make sense since the data spans such a wide time period 1922 to 2017. The NA distinction just doesn't tell us anything in the context of our question.
```{r}
# assess what projects have the most salmon counts
sampled_salmon |> 
  filter(aslprojecttype != "999") |> # remove any NA projects from the investigation
  group_by(aslprojecttype) |> 
  summarize(sum = sum(num_sample)) |> 
  arrange(desc(sum))
```

The top three project types are commercial catch, escapement, and hatchery cost recovery. 

### Create a color palette

This color palette will be used in the visual below. Though some may argue that color is unnecessary for this visualization since there is one y-axis furthest left, the color gives an easy way to interpret what species each bar is related to.
```{r}
pal <- c("chum" = "#FF9F8C",
         "chinook" = "#FF8B55", 
         "pink" = "#CD5B47",
         "sockeye" = "#913E3E",
         "coho" = "#2E2824")
```


Again top three ASL project type is commercial catch, escapement, and hatchery cost recovery. Let's make a visualization showing the count by salmon species!

```{r}

sampled_salmon |> 
  filter(aslprojecttype == "commercial catch" 
         | aslprojecttype == "escapement" 
         | aslprojecttype == "hatchery cost recovery") |>  # select for the project types of interes
  ggplot() +
  geom_col(aes(y = num_sample,
               x = reorder(species, -num_sample),  # arrange by count of each species
               fill = species),
           show.legend = FALSE) +
  scale_fill_manual(values = pal) +
  facet_wrap(vars(aslprojecttype)) + 
  coord_flip() +
  theme_minimal() + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) + # maintain a raw number axis
  labs(x = "Salmon species",
       y = "Count") 
```
Sockeye salmon are the salmon species that has been most caught through commercial catch and escapement. 

#### A closer look at hatchery cost recovery

The hatchery cost recovery counts per species are a lot less than the count for commercial catch and escapement. These values are hard to see in the faceted plot. Here we create a plot to scale based on only the hatchery cost recovery count numbers. 

```{r}
sampled_salmon |> 
  filter(aslprojecttype == "hatchery cost recovery") |> 
  ggplot() +
  geom_col(aes(y = num_sample, 
               x = reorder(species, -num_sample), 
               fill = species),
           show.legend = FALSE) +
  scale_fill_manual(values = pal) +
  coord_flip() +
  theme_minimal() + 
  labs(x = "Salmon species",
       y = "Count",
       title = "Hatchery Cost Recovery")
```
Chum salmon are the salmon species that have been caught most often through hatchery catch recovery.


## Close the database connection

```{r}
DBI::dbDisconnect(conn, shutdown = TRUE)
```

